---
title: Assess topic model fits in newsgroups data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM and SCD, with and without extrapolation), and with
different numbers of topics, $K$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in the analysis below, as well as some
additional functions for creating the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(ggrepel)
set.seed(1)
source("../code/plot_functions.R")
```

Load the newsgroups data, and the results of running `fit_poisson_nmf`
on the "20 newsgroups" data, with different algorithms, and for
various settings of $K$. Also load the results of running LDA with
different initializations.

```{r load-data}
load("../data/newsgroups.RData")
load("../output/newsgroups/fits-newsgroups.RData")
fits <- lapply(fits,poisson2multinom)
```

This plot shows the improvement in the log-likelihood as the rank,
$K$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $K = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=3.5}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

The next set of plots shows the improvement in the fit over time, for
$K$ from 2 to 12, using EM or SCD, with and without extrapolation. The
quality of the fit is measured by the log-likelihood relative to the
best log-likelihood that was identified among all methods compared.

```{r plot-loglik, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
prune_prefit_iters <- function (fit) {
  n <- nrow(fit$progress)
  fit$progress <- fit$progress[1000:n,]
  fit$progress <- transform(fit$progress,timing = timing/60^2)
  return(fit)
}
create_progress_plot <- function (fits, k, y = "loglik")
  plot_progress(fits,y = y,add.point.every = 100,shapes = 21,
                colors = c("dodgerblue","red","dodgerblue","red"),
                fills = c("dodgerblue","red","white","white")) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-8,8)) +
  guides(color = "none",fill = "none",size = "none",
         shape = "none",linetype = "none") +
  labs(x = "runtime (h)",title = paste("K =",k)) +
  theme_cowplot(font_size = 10) +
  theme(plot.title = element_text(size = 10,face = "plain"))
fits <- lapply(fits,prune_prefit_iters)
p <- vector("list",12)
for (i in 2:12)
  p[[i]] <- create_progress_plot(fits[dat$k == i],i)
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-1, message=FALSE, warning=FALSE, echo=FALSE}
ggsave("../plots/progress_loglik_newsgroups.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```

These plots shows the evolution of the KKT residuals over time.

```{r plot-res, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
for (i in 2:12)
  p[[i]] <- create_progress_plot(fits[dat$k == i],i,y = "res")
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-2, message=FALSE, warning=FALSE, echo=FALSE}
ggsave("../plots/progress_kkt_newsgroups.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```

An example in which the EM and (extrapolated) CD estimates largely
agree:

```{r loadings-scatterplot-k-9, fig.width=3.5, fig.height=3}
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
                  "#e31a1c","#fdbf6f","#ff7f00","#cab2d6","#6a3d9a")
fit1 <- fits[["fit-newsgroups-em-k=9"]]
fit2 <- fits[["fit-newsgroups-scd-ex-k=9"]]
n    <- nrow(fit1$L)
pdat <- data.frame(x = as.vector(fit1$L),
                   y = as.vector(fit2$L),
				   k = factor(rep(1:9,each = n)))
pdat <- pdat[sample(9*n),]
p1 <- ggplot(pdat,aes(x = x,y = y,fill = k)) +
  geom_point(color = "white",shape = 21,size = 2) +
  geom_abline(color = "black",linetype = "dotted") +
  scale_fill_manual(values = topic_colors) +
  labs(x = "EM estimate",y = "extrapolated CD estimate") +
  theme_cowplot(font_size = 12)
print(p1)
```

An example in which the EM and (extrapolated) CD estimates greatly
differ:

```{r loadings-scatterplot-k-10, fig.width=3.75, fig.height=3}
fit1 <- fits[["fit-newsgroups-em-k=10"]]
fit2 <- fits[["fit-newsgroups-scd-ex-k=10"]]
n    <- nrow(fit1$L)
pdat <- data.frame(x = as.vector(fit1$L),
                   y = as.vector(fit2$L),
				   k = factor(rep(1:10,each = n)))
pdat <- pdat[sample(10*n),]
p2 <- ggplot(pdat,aes(x = x,y = y,fill = k)) +
  geom_point(color = "white",shape = 21,size = 2) +
  geom_abline(color = "black",linetype = "dotted") +
  scale_fill_manual(values = topic_colors) +
  labs(x = "EM estimate",y = "extrapolated CD estimate") +
  theme_cowplot(font_size = 12)
print(p2)
```

For a closer look at the $K = 10$ (extrapolated) CD estimates, we
create a Structure plot:

```{r, fig.width=8, fig.height=2, message=FALSE}
set.seed(1)
topics <- factor(topics,
                 c("sci.med","rec.autos","rec.motorcycles","alt.atheism",
				 "soc.religion.christian","talk.religion.misc",
				 "rec.sport.baseball","rec.sport.hockey",
				 "talk.politics.mideast","talk.politics.guns","sci.crypt",
				 "talk.politics.misc","sci.space","sci.electronics",
				 "misc.forsale","comp.sys.ibm.pc.hardware",
				 "comp.sys.mac.hardware","comp.os.ms-windows.misc",
				 "comp.graphics","comp.windows.x"))
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
                  "#e31a1c","#fdbf6f","#ff7f00","#cab2d6","#6a3d9a")
fit <- fits[["fit-newsgroups-scd-ex-k=10"]]
p3 <- structure_plot(fit,grouping = topics,topics = 10:1,
                     colors = topic_colors[10:1],perplexity = 20,gap = 20,
					 verbose = FALSE)
print(p3)
```

Topic 1 appears to be capturing "off-topic" discussion:

```{r words-scatterplot, fig.width=4.25, fig.height=4}
pdat <- data.frame(word = colnames(counts),
                   x    = pmax(1e-6,apply(fit$F[,-1],1,max)),
                   y    = pmax(1e-6,fit$F[,1]),
				   stringsAsFactors = FALSE)
pdat <- transform(pdat,fc = y/x)
pdat[with(pdat,!((fc > 4 & x > 1e-5) | (fc > 2 & x > 0.002))),"word"] <- ""
p4 <- ggplot(pdat,aes(x = x,y = y,label = word)) +
  geom_point(color = "white",fill = "royalblue",shape = 21) +
  geom_text_repel(color = "darkgray",size = 2.25,segment.color = "darkgray",
                  segment.size = 0.25,min.segment.length = 0,
				  max.overlaps = Inf) +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_x_continuous(trans = "log10",limits=10^c(-6,-1),breaks=10^seq(-6,-1)) +
  scale_y_continuous(trans = "log10",limits=10^c(-6,-1),breaks=10^seq(-6,-1)) +
  labs(x = "highest frequency in another topic",
       y = "frequency in topic 1") +
  theme_cowplot(font_size = 10)
print(p4)
```

```{r plots-for-paper-3, message=FALSE, echo=FALSE}
ggsave("../plots/loadings_scatterplot_newsgroups_k9.eps",
       p1,height = 2.5,width = 3)
ggsave("../plots/loadings_scatterplot_newsgroups_k10.eps",
       p2,height = 2.5,width = 3.25)
ggsave("../plots/structure_plot_newsgroups.eps",p3,height = 2,width = 8)
ggsave("../plots/words_scatterplot_newsgroups.eps",p4,height = 4,width = 4.25)
```

Finally, let's have a look at the results of running LDA with the
various initializations:

```{r elbo-plots, fig.height=6, fig.width=9}
load("../output/newsgroups/lda-newsgroups.RData")
p <- vector("list",12)
for (i in 2:12) {
  runs   <- which(dat$k == i)
  p[[i]] <- create_elbo_plot(fits[runs],dat[runs,"runtime"],i)
}
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-4, message=FALSE, echo=FALSE}
ggsave("../plots/lda_elbo_newsgroups.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```
