---
title: Assess topic model fits in NIPS data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM and SCD, with and without extrapolation), and with
different numbers of topics, $K$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in the analysis below, as well as some
additional functions for creating the plots.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Load the results of running `fit_poisson_nmf` on the NIPS data,
with different algorithms, and for various settings of $K$.

```{r load-results}
load("../output/nips/fits-nips.RData")
```

This plot shows the improvement in the log-likelihood as the rank,
$K$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $K = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=5, eval=FALSE}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

Oddly, the likelihood did not improve from 4 to 5, and same from 8
to 9. Otherwise, the likelihood improved with larger rank, $K$.

The next set of plots shows the improvement in the fit over time, for
$K$ from 2 to 13, and for each of the three updates (EM, CCD, SCD)
with and without extrapolation. The quality of the fit is measured by
the log-likelihood relative to the best log-likelihood that was
identified among all methods compared.

```{r plot-loglik, fig.height=6, fig.width=9, eval=FALSE}
create_progress_plots(dat,fits,"loglik")
```

The last set of plots shows the evolution of the KKT residuals over
time; the KKT residuals should vanish near a stationary point of the
log-likelihood, so looking at the largest KKT residual can be used to
assess how close we are to a solution. Note that, unlike the
log-likelihood, the KKT residuals are not expected to decrease
monotonically over time.

```{r plot-res, fig.height=6, fig.width=9, eval=FALSE}
create_progress_plots(dat,fits,"res")
```
