---
title: Assess topic model fits in 68k PBMC data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM and SCD, with and without extrapolation), and with
different numbers of topics, $K$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in the analysis below, as well as some
additional functions for creating the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(ggrepel)
set.seed(1)
```

Load the 68k PBMC data, and the results of running `fit_poisson_nmf`
on the 68k PBMC data, with different algorithms, and for various
settings of $K$.

```{r load-results}
load("../data/pbmc_68k.RData")
load("../output/pbmc68k/fits-pbmc68k.RData")
fits <- lapply(fits,poisson2multinom)
```

This plot shows the improvement in the log-likelihood as the rank,
$K$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $K = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=3.5}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

The next set of plots shows the improvement in the fit over time, for
$K$ from 2 to 12, using EM or SCD, with and without extrapolation. The
quality of the fit is measured by the log-likelihood relative to the
best log-likelihood that was identified among all methods compared.

```{r plot-loglik, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
prune_prefit_iters <- function (fit) {
  n <- nrow(fit$progress)
  fit$progress <- fit$progress[1000:n,]
  fit$progress <- transform(fit$progress,timing = timing/60^2)
  return(fit)
}
create_progress_plot <- function (fits, k, y = "loglik")
  plot_progress(fits,y = y,add.point.every = 100,shapes = 21,
                colors = c("dodgerblue","red","dodgerblue","red"),
                fills = c("dodgerblue","red","white","white")) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-8,8)) +
  guides(color = "none",fill = "none",size = "none",
         shape = "none",linetype = "none") +
  labs(x = "runtime (h)",title = paste("K =",k)) +
  theme_cowplot(font_size = 10) +
  theme(plot.title = element_text(size = 10,face = "plain"))
fits <- lapply(fits,prune_prefit_iters)
p <- vector("list",12)
for (i in 2:12)
  p[[i]] <- create_progress_plot(fits[dat$k == i],i)
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-1, message=FALSE, warning=FALSE, echo=FALSE}
ggsave("../plots/progress_loglik_pbmc68k.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```

These plots shows the evolution of the KKT residuals over time.

```{r plot-res, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
for (i in 2:12)
  p[[i]] <- create_progress_plot(fits[dat$k == i],i,y = "res")
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-2, message=FALSE, warning=FALSE, echo=FALSE}
ggsave("../plots/progress_kkt_pbmc68k.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```

An example in which the EM and (extrapolated) CD estimates greatly
differ:

```{r loadings-scatterplot-k-7, fig.width=3.75, fig.height=3}
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c",
                  "#fb9a99","#e31a1c","#fdbf6f")
fit1 <- fits[["fit-pbmc68k-em-k=7"]]
fit2 <- fits[["fit-pbmc68k-scd-ex-k=7"]]
n    <- nrow(fit1$L)
pdat <- data.frame(x = as.vector(fit1$L),
                   y = as.vector(fit2$L),
				   k = factor(rep(1:7,each = n)))
pdat <- pdat[sample(7*n),]
p1 <- ggplot(pdat,aes(x = x,y = y,fill = k)) +
  geom_point(color = "white",shape = 21,size = 2) +
  scale_fill_manual(values = topic_colors) +
  labs(x = "EM estimate",y = "extrapolated CD estimate") +
  theme_cowplot(font_size = 12)
print(p1 + geom_abline(color = "black",linetype = "dotted"))
```

An example in which the EM and (extrapolated) CD estimates largely
agree:

```{r loadings-scatterplot-k-12, fig.width=3.75, fig.height=3}
topic_colors <- c("#8dd3c7","#ffffb3","#bebada","#fb8072","#80b1d3","#fdb462",
                  "#b3de69","#fccde5","#d9d9d9","#bc80bd","#ccebc5","#ffed6f")
fit1 <- fits[["fit-pbmc68k-em-k=12"]]
fit2 <- fits[["fit-pbmc68k-scd-ex-k=12"]]
n    <- nrow(fit1$L)
pdat <- data.frame(x = as.vector(fit1$L),
                   y = as.vector(fit2$L),
				   k = factor(rep(1:12,each = n)))
pdat <- pdat[sample(12*n),]
p2 <- ggplot(pdat,aes(x = x,y = y,fill = k)) +
  geom_point(color = "white",shape = 21,size = 2) +
  scale_fill_manual(values = topic_colors) +
  labs(x = "EM estimate",y = "extrapolated CD estimate") +
  theme_cowplot(font_size = 12)
print(p2 + geom_abline(color = "black",linetype = "dotted"))
```

From the PCs of the mixture proportions, we define four clusters,
roughly corresponding to (1) B cells, (2) dendritic and CD14+
monocytes, (3) CD34+ cells, and (4) T cells and natural killer cells:

```{r clustering}
fit  <- fits[["fit-pbmc68k-scd-ex-k=7"]]
pca  <- prcomp(fit$L)$x
n    <- nrow(pca)
x    <- rep("T+NK",n)
pc4  <- pca[,4]
pc6  <- pca[,6]
x[pc4 > 0.25] <- "B"
x[pc4 < -0.25] <- "CD14+dendritic"
x[pc6 > 0.8] <- "CD34+"
clusters <- factor(x)
table(clusters)
```

The Structure plot summarizes the topic proportions in the cells. Here
we subsample the larger clusters in order to better visualize the
rarer cell types.

```{r structure-plot, fig.width=8, fig.height=2, message=FALSE}
set.seed(1)
topics <- c(1:5,7,6)
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c",
                  "#fb9a99","#e31a1c","#fdbf6f")
rows <- sort(c(sample(which(clusters == "B"),746),
               sample(which(clusters == "CD14+dendritic"),806),
               which(clusters == "CD34+"),
               sample(which(clusters == "T+NK"),1213)))
p3 <- structure_plot(select_loadings(fit,loadings = rows),
                     grouping = clusters[rows],topics = topics,
                     colors = topic_colors[topics],n = Inf,perplexity = 70,
                     gap = 30,num_threads = 4,verbose = FALSE)
print(p3)					 
```

Many of the "driving genes" in topic 6---that is, genes showing the
largest increase in expression relative to other topics---are
ribosomal protein genes:

```{r words-scatterplot, fig.width=4.25, fig.height=4}
pdat <- data.frame(gene = genes$symbol,
                   x    = pmax(1e-6,apply(fit$F[,-6],1,max)),
                   y    = pmax(1e-6,fit$F[,6]),
				   stringsAsFactors = FALSE)
pdat <- transform(pdat,fc = y/x)
pdat[with(pdat,!(fc > 1.6 & x > 0.0005)),"gene"] <- ""
p4 <- ggplot(pdat,aes(x = x,y = y,label = gene)) +
  geom_point(color = "white",fill = "royalblue",shape = 21) +
  geom_text_repel(color = "darkgray",size = 2.25,segment.color = "darkgray",
                  segment.size = 0.25,min.segment.length = 0,
				  max.overlaps = Inf) +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_x_continuous(trans = "log10",limits=10^c(-6,-1),breaks=10^seq(-6,-1)) +
  scale_y_continuous(trans = "log10",limits=10^c(-6,-1),breaks=10^seq(-6,-1)) +
  labs(x = "highest frequency in another topic",
       y = "frequency in topic 6") +
  theme_cowplot(font_size = 10)
print(p4)
```

```{r plots-for-paper-3, message=FALSE, echo=FALSE}
ggsave("../plots/loadings_scatterplot_pbmc68k_k7.png",
       p1,height = 2.5,width = 3,dpi = 900)
ggsave("../plots/loadings_scatterplot_pbmc68k_k12.png",
       p2,height = 2.5,width = 3,dpi = 900)
ggsave("../plots/structure_plot_pbmc68k.eps",p3,height = 2,width = 8)
ggsave("../plots/genes_scatterplot_pbmc68k.eps",p4,height = 4,width = 4.25)
```
