---
title: Assess topic model fits in NIPS data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM and SCD, with and without extrapolation), and with
different numbers of topics, $K$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in the analysis below, as well as some
additional functions for creating the plots.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Load the results of running `fit_poisson_nmf` on the NIPS data,
with different algorithms, and for various settings of $K$.

```{r load-results}
load("../output/nips/fits-nips.RData")
fits <- lapply(fits,poisson2multinom)
```

This plot shows the improvement in the log-likelihood as the rank,
$K$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $K = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=5}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

The next set of plots shows the improvement in the fit over time, for
$K$ from 2 to 12, using EM or SCD, with and without extrapolation. The
quality of the fit is measured by the log-likelihood relative to the
best log-likelihood that was identified among all methods compared.

```{r plot-loglik, fig.height=6, fig.width=9, eval=FALSE}
prune_prefit_iters <- function (fit) {
  n <- nrow(fit$progress)
  fit$progress <- fit$progress[1000:n,] 
  return(fit)
}
create_progress_plot <- function (fits)
  plot_progress(fits,add.point.every = 100,shapes = 21,
                colors = c("dodgerblue","red","dodgerblue","red"),
                fills = c("dodgerblue","red","white","white")) +
  guides(color = "none",fill = "none",size = "none",
         shape = "none",linetype = "none") +
  theme_cowplot(font_size = 10) +
  labs(x = "runtime (h)")
fits <- lapply(fits,prune_prefit_iters)
p2  <- create_progress_plot(fits[dat$k == 2])
p3  <- create_progress_plot(fits[dat$k == 3])
p4  <- create_progress_plot(fits[dat$k == 4])
p5  <- create_progress_plot(fits[dat$k == 5])
p6  <- create_progress_plot(fits[dat$k == 6])
p7  <- create_progress_plot(fits[dat$k == 7])
p8  <- create_progress_plot(fits[dat$k == 8])
p9  <- create_progress_plot(fits[dat$k == 9])
p10 <- create_progress_plot(fits[dat$k == 10])
p11 <- create_progress_plot(fits[dat$k == 11])
p12 <- create_progress_plot(fits[dat$k == 12])
plot_grid(p2,p3,p4,p5,
          p6,p7,p8,p9,
          p10,p11,p12,
          nrow = 3,ncol = 4)
```

The last set of plots shows the evolution of the KKT residuals over
time.

```{r plot-res, fig.height=6, fig.width=9, eval=FALSE}
```
