---
title: Assess topic model fits in NIPS data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM and SCD, with and without extrapolation), and with
different numbers of topics, $K$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in the analysis below, as well as some
additional functions for creating the plots.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Load the results of running `fit_poisson_nmf` on the NIPS data,
with different algorithms, and for various settings of $K$.

```{r load-results}
load("../output/nips/fits-nips.RData")
fits <- lapply(fits,poisson2multinom)
```

This plot shows the improvement in the log-likelihood as the rank,
$K$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $K = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=5}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

The next set of plots shows the improvement in the fit over time, for
$K$ from 2 to 12, using EM or SCD, with and without extrapolation. The
quality of the fit is measured by the log-likelihood relative to the
best log-likelihood that was identified among all methods compared.

```{r plot-loglik, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
prune_prefit_iters <- function (fit) {
  n <- nrow(fit$progress)
  fit$progress <- fit$progress[1000:n,]
  return(fit)
}
create_progress_plot <- function (fits, k, y = "loglik")
  plot_progress(fits,y = y,add.point.every = 100,shapes = 21,
                colors = c("dodgerblue","red","dodgerblue","red"),
                fills = c("dodgerblue","red","white","white")) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-8,8)) +
  guides(color = "none",fill = "none",size = "none",
         shape = "none",linetype = "none") +
  labs(x = "runtime (s)",title = paste("K =",k)) +
  theme_cowplot(font_size = 10) +
  theme(plot.title = element_text(size = 10,face = "plain"))
fits <- lapply(fits,prune_prefit_iters)
p <- vector("list",12)
for (i in 2:12)
  p[[i]] <- create_progress_plot(fits[dat$k == i],i)
p[[2]] <- p[[2]] + scale_y_continuous()
p[[3]] <- p[[3]] + scale_y_continuous()
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-1, message=FALSE, warning=FALSE, echo=FALSE}
ggsave("../plots/progress_loglik_nips.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```

These plots shows the evolution of the KKT residuals over time.

```{r plot-res, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
for (i in 2:12)
  p[[i]] <- create_progress_plot(fits[dat$k == i],i,y = "res")
plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
          p[[6]],p[[7]],p[[8]],p[[9]],
          p[[10]],p[[11]],p[[12]],
          nrow = 3,ncol = 4)
```

```{r plots-for-paper-2, message=FALSE, warning=FALSE, echo=FALSE}
ggsave("../plots/progress_kkt_nips.eps",
       plot_grid(p[[2]],p[[3]],p[[4]],p[[5]],
                 p[[6]],p[[7]],p[[8]],p[[9]],
                 p[[10]],p[[11]],p[[12]],
                 nrow = 3,ncol = 4),
   	   height = 6,width = 9)
```

For the most part, the EM and CD algorithms achieve similar estimates
in this data set. For example, for $K = 10$, the difference in the
topic model likelihoods between the EM and CD estimates is very small,
and indeed the estimated topic proportions are nearly identical:

```{r loadings-scatterplot, fig.width=3, fig.height=3}
fit1 <- fits[["fit-nips-em-k=10"]]
fit2 <- fits[["fit-nips-scd-k=10"]]
pdat <- data.frame(x = as.vector(fit1$L),y = as.vector(fit2$L))
p1 <- ggplot(pdat,aes(x = x,y = y)) +
  geom_point(shape = 21,size = 2,color = "white",fill = "royalblue") +
  geom_abline(color = "black",linetype = "dotted") +
  labs(x = "EM estimate",y = "CD estimate") +
  theme_cowplot(font_size = 12)
print(p1)
```

```{r plots-for-paper-3, message=FALSE, echo=FALSE}
ggsave("../plots/loadings_scatterplot_nips_k10.eps",
       p1,height = 2.5,width = 2.6)
```
