---
title: Find keywords in newsgroups topics
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we run the "differential expression" (DE) analysis on the $K =
10$ topic model fitted to the newsgroups data to identify topic
keywords.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Load the newsgroups data, the $K = 10$ topic model fit, and the
results of running the DE analysis with this topic model.

```{r load-data}
load("../data/newsgroups.RData")
load("../output/newsgroups/fits-newsgroups.RData")
load("../output/newsgroups/de-newsgroups.RData")
fit <- fits[["fit-newsgroups-scd-ex-k=10"]]
fit <- poisson2multinom(fit)
```

Create a Structure plot from the topic model.

```{r structure-plot, fig.height=2.25, fig.width=9, message=FALSE}
topics <- factor(topics,
                 c("sci.med","rec.autos","rec.motorcycles","alt.atheism",
                 "soc.religion.christian","talk.religion.misc",
                 "rec.sport.baseball","rec.sport.hockey",
                 "talk.politics.mideast","talk.politics.guns","sci.crypt",
                 "talk.politics.misc","sci.space","sci.electronics",
                 "misc.forsale","comp.sys.ibm.pc.hardware",
                 "comp.sys.mac.hardware","comp.os.ms-windows.misc",
                 "comp.graphics","comp.windows.x"))
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
                  "#e31a1c","#fdbf6f","#ff7f00","#cab2d6","#6a3d9a")
structure_plot(fit,grouping = topics,topics = 10:1,colors = topic_colors,
               perplexity = 20,gap = 20,verbose = FALSE)
```

Next, we visualize the keywords for topic 2 (dark blue) in a volcano
plot. From the top keywords (highest LFC), this topic seems to be
discussion about cryptography and privacy.

```{r volcano-plot-topic-2, fig.height=3.5, fig.width=4, message=FALSE}
volcano_plot(de,k = 2,ymax = 25,
             do.label = function (lfc, z) lfc > 5 | (z > 10 & lfc > 2)) +
  scale_y_continuous(trans = "identity")
```

Topic 7 (light orange) appears to be discussion of health and
medicine:

```{r volcano-plot-topic-7, fig.height=3.5, fig.width=4, message=FALSE}
volcano_plot(de,k = 7,ymax = 20,
             do.label = function (lfc, z) lfc > 5 | (z > 10 & lfc > 2)) +
  scale_y_continuous(trans = "identity")
```

Topic 9 (light purple) is about sports:

```{r volcano-plot-topic-9, fig.height=3.5, fig.width=4, message=FALSE}
volcano_plot(de,k = 9,ymax = 25,
             do.label = function (lfc, z) lfc > 6 | (z > 12 & lfc > 2)) +
  scale_y_continuous(trans = "identity")
```

Topic 1 contributes to all the newsgroup discussions, and appears to
be some sort of "catch all" topic for words that are not
newsgroup-specific:

```{r volcano-plot-topic-1, fig.height=3.5, fig.width=4, message=FALSE}
volcano_plot(de,k = 1,ymax = 30,
             do.label = function (lfc, z) lfc > 2 | (z > 10 & lfc > 1)) +
  scale_y_continuous(trans = "identity")
```
