---
title: EM vs. SCD on the MCF-7 data set
author: Peter Carbonetto
output: workflowr::wflow_html
---

This will be the new in-depth example for the paper illustrating the
differences in performance of the EM and SCD methods. See
[here](https://stephenslab.github.io/single-cell-jamboree/mcf7.html)
for background on the MCF-7 data set, including steps taken to prepare
the data.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First load the packages used in the code below:

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Set the seed to ensure that the results are reproducible:

```{r set-seed}
set.seed(1)
```

Load the MCF-7 data set and initial estimate of the topic proportions:

```{r load-data}
load("../data/mcf7.RData")
dim(counts)
```

I obtain a "smart" initialization by running 4 EM iterations:

```{r initialize-topic-model}
L <- L[,2:4]
control <- list(extrapolate = FALSE,numiter = 4,nc = 8)
fit0 <- init_poisson_nmf(counts,L = L,init.method = "random")
fit0 <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 4,method = "em",
                        control = control,verbose = "none")
```

This "smart" initialization will be used as the starting point for
all the comparisons.

This is what the smart initialization looks like:

```{r structure-plot-initial, fig.height=1.25, fig.width=4}
topic_colors <- c("darkblue","dodgerblue","tomato")
n <- nrow(samples)
p1 <- structure_plot(fit0,grouping = samples$label,topics = c(3,1,2),
                     colors = topic_colors,loadings_order = 1:n) +
  labs(y = "topic prop.") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
print(p1)
```

Now let's run the extrapolated SCD algorithm for a decently long time
to obtain a high-quality fit:

```{r fit-poisson-nmf-high-quality, cache=TRUE}
control$extrapolate <- FALSE
fit_best <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 100,method = "scd",
                            control = control,verbose = "none")
control$extrapolate <- TRUE
fit_best <- fit_poisson_nmf(counts,fit0 = fit_best,numiter = 200,
                            method = "scd",control = control,
							verbose = "none")
```

These are the max. residuals in the last iterations:

```{r check-high-quality-fit}
tail(fit_best$progress$res,n = 4)
```

Here's what the topic model looks like:

```{r structure-plot-high-quality, fig.height=1.25, fig.width=4}
p2 <- structure_plot(fit_best,grouping = samples$label,topics = c(3,1,2),
                     colors = topic_colors,loadings_order = 1:n) +
  labs(y = "topic prop.") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
print(p2)
```

Fit the topic model by performing EM updates or SCD updates, with or
without extrapolation.

```{r fit-poisson-nmf-more, cache=TRUE}
control$extrapolate <- FALSE
fit_em  <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 100,method = "em",
                           control = control,verbose = "none")
fit_scd <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 200,method = "scd",
                           control = control,verbose = "none")
fit_scd_ex <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 50,method = "scd",
                              control = control,verbose = "none")
control$extrapolate <- TRUE
fit_scd_ex <- fit_poisson_nmf(counts,fit0 = fit_scd_ex,numiter = 50,
                              method = "scd",control = control,
							  verbose = "none")
fit_em_scd <- fit_poisson_nmf(counts,fit0 = fit_em,numiter = 100,
                              method = "scd",control = control,
							  verbose = "none")
control$extrapolate <- FALSE
fit_em  <- fit_poisson_nmf(counts,fit0 = fit_em,numiter = 100,method = "em",
                           control = control,verbose = "none")
```

This plot shows the improvement in the solution over time for the
different methods:

```{r progress-plots, fig.height=2.5, fig.width=4}
progress_plot_colors <- c("dodgerblue","darkblue","darkorange","magenta")
ll_best <- tail(fit_best$progress$loglik.multinom,n = 1)
pdat <- rbind(cbind(fit_em$progress,data.frame(method = "em")),
              cbind(fit_scd$progress,data.frame(method = "scd")),
			  cbind(fit_scd_ex$progress,data.frame(method = "scd+ex")),
			  cbind(fit_em_scd$progress,data.frame(method = "em+scd+ex")))
pdat <- transform(pdat,
                  loglik.multinom = ll_best - loglik.multinom,
                  method = factor(method))
pdat2 <- subset(pdat,iter %% 50 == 1)
p3 <- ggplot(pdat,aes(x = iter,y = loglik.multinom,color = method)) +
  geom_line() +
  geom_point(data = pdat2,size = 1) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-3,7)) +
  scale_color_manual(values = progress_plot_colors) +
  labs(x = "iteration",y = "distance from best loglik") +
  theme_cowplot(font_size = 12)
print(p3)
```

Compare the log-likelihoods obtained by the different methods
(relative to the log-likelihood at the "high-quality" solution):

```{r logliks}
logliks <- c("initial"   = sum(loglik_multinom_topic_model(counts,fit0)),
             "em"        = sum(loglik_multinom_topic_model(counts,fit_em)),
             "scd"       = sum(loglik_multinom_topic_model(counts,fit_scd)),
             "scd+ex"    = sum(loglik_multinom_topic_model(counts,fit_scd_ex)),
             "em+scd+ex" = sum(loglik_multinom_topic_model(counts,fit_em_scd)))
ll_best - logliks
```

Compare the fit obtained by each method in Structure plots:

```{r more-structure-plots, fig.width=4, fig.height=5.5}
p4 <- structure_plot(fit_em,grouping = samples$label,topics = c(3,1,2),
                    colors = topic_colors,loadings_order = 1:n) +
  labs(y = "topic prop.",title = "EM") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
p5 <- structure_plot(fit_scd,grouping = samples$label,topics = c(3,1,2),
                     colors = topic_colors,loadings_order = 1:n) +
  labs(y = "topic prop.",title = "SCD") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
p6 <- structure_plot(fit_scd_ex,grouping = samples$label,topics = c(3,1,2),
                     colors = topic_colors) +
  labs(y = "topic prop.",title = "SCD + extrapolate") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
p7 <- structure_plot(fit_em_scd,grouping = samples$label,topics = c(3,1,2),
                     colors = topic_colors) +
  labs(y = "topic prop.",title = "EM + SCD + extrapolate") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
print(plot_grid(p4,p5,p6,p7,nrow = 4,ncol = 1))
```

Also, let's check that the EM updates *eventually* converge to the
same fixed point:

```{r fit-poisson-nmf-em-long, cache=TRUE}
control$extrapolate <- FALSE
fit_em_long  <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 4000,
                                method = "em",control = control,
                                verbose = "none")
```

Indeed, they do:

```{r check-em-long}
cor(poisson2multinom(fit_best)$L,poisson2multinom(fit_em_long)$L)
```

Plot the improvement in the EM updates over time:

```{r progress-plot, fig.height=2, fig.width=3}
pdat <- fit_em_long$progress
pdat <- transform(pdat,loglik.multinom = ll_best - loglik.multinom)
pdat2 <- subset(pdat,iter %% 50 == 1)
p8 <- ggplot(pdat,aes(x = iter,y = loglik.multinom)) +
  geom_line(color = "darkorange") +
  geom_point(data = pdat2,size = 1,color = "darkorange") +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-2,+7)) +
  labs(x = "iteration",y = "dist. from best loglik") +
  theme_cowplot(font_size = 12)
print(p8)
```

Finally, save the key plots to PDF:

```{r ggsave}
ggsave("mcf7_structure_plots.pdf",
       plot_grid(p1 + ggtitle("initial"),
	             p2 + ggtitle("solution"),
				 p4,
				 p5,
				 nrow = 4,ncol = 1),
	   height = 5.5,width = 4)
ggsave("mcf7_progress_plot.pdf",p3,height = 2.5,width = 4)
ggsave("mcf7_progress_plot_em.pdf",p8,height = 2,width = 3)
```
