---
title: A closer look at some results on the newsgroups data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we take a closer look at some of the results on the newsgroups data.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(topicmodels)
library(fastTopics)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Load the newsgroups data.

```{r load-data}
load("../data/newsgroups.RData")
```

Load the topic models fit using the EM and CD algorithms

```{r load-fits}
fit1 <- readRDS("../output/newsgroups/rds/fit-newsgroups-em-k=10.rds")$fit
fit2 <- readRDS("../output/newsgroups/rds/fit-newsgroups-scd-ex-k=10.rds")$fit
fit1 <- poisson2multinom(fit1)
fit2 <- poisson2multinom(fit2)
```

and the LDA fits initialized using the EM and CD estimates:

```{r load-lda}
lda1 <- readRDS("../output/newsgroups/rds/lda-newsgroups-em-k=10.rds")$lda
lda2 <- readRDS("../output/newsgroups/rds/lda-newsgroups-scd-ex-k=10.rds")$lda
```

The MLEs and the approximate posterior estimates from LDA turn out to
be very similar to each other, so there is really no need to examine
both. Here we'll focus on the MLEs:

```{r compare-mle-vs-lda}
cor(as.vector(fit1$L),as.vector(lda1@gamma))
cor(as.vector(fit2$L),as.vector(lda2@gamma))
```

Let's now examine the results using Structure plots. Here are the EM
estimates:

```{r structure-plot-em, fig.height=2.25, fig.width=8, results="hide", message=FALSE} 
n    <- nrow(fit1$L)
rows <- sample(n,2000)
L1   <- select_loadings(fit1,rows)$L
topics <- factor(topics,
                 c("rec.sport.hockey",
                   "rec.sport.baseball",
                   "sci.med",
                   "comp.graphics",
                   "comp.windows.x",
                   "comp.os.ms-windows.misc",
                   "comp.sys.ibm.pc.hardware",
                   "comp.sys.mac.hardware",
                   "misc.forsale",
                   "sci.electronics",
                   "sci.space",
                   "alt.atheism",
                   "soc.religion.christian",
                   "talk.religion.misc",
                   "rec.autos",
                   "rec.motorcycles",
                   "sci.crypt",
                   "talk.politics.misc",
                   "talk.politics.guns",
                   "talk.politics.mideast"))
topic_ordering <- c(2:10,1)
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
                  "#e31a1c","#fdbf6f","#ff7f00","#cab2d6","#6a3d9a")
p1 <- structure_plot(L1,topics = 1:10,grouping = topics[rows],
                     colors = topic_colors,gap = 20) +
  ggtitle("EM without extrapolation") +
  theme(plot.title = element_text(face = "plain",size = 10))
p1
```

And here are the CD estimates:

```{r structure-plot-cd, fig.height=2.25, fig.width=8, results="hide", message=FALSE}
L2 <- select_loadings(fit2,rows)$L
p2 <- structure_plot(L2,topics = 1:10,grouping = topics[rows],
                     colors = topic_colors,gap = 20) +
  ggtitle("CD with extrapolation") +
  theme(plot.title = element_text(face = "plain",size = 10))
p2
```

The most striking differences are in topics 1 and 8.

Let's now extract some "keywords" for a few selected topics by taking
words that are at higher frequency in the given topic compared to the
other topics. For example, top keywords for topic 9 clearly relate to
baseball, hockey and sports more generally:

```{r keywords-topic-9}
k <- 9
dat <- data.frame(word = colnames(counts),
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
subset(dat,f0 < 1e-5 & f2 > 1e-3)
```

The keywords for topic 1 seem to suggest a "background topic" that
captures words that are not specific to any topic:

```{r keywords-topic-1}
k <- 1
dat <- data.frame(word = colnames(counts),
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
subset(dat,f0 > 1e-6 & f2/f0 > 5)
```

Finally, topic 8 is a topic that is quite noticeably different between
the EM and CD estimates, and indeed based on the keywords, only the CD
estimates produce a topic about cars and motorcycles, with keywords
such as wheel, riding, bmw, etc:

```{r keywords-topic-8}
k <- 8
dat <- data.frame(word = colnames(counts),
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
subset(dat,f0 < 1e-5 & f2 > 5e-4)
```

```{r plots-for-paper, echo=FALSE}
ggsave("../plots/structure_plots_newsgroups.pdf",
       plot_grid(p1,p2,nrow = 2,ncol = 1),
	   height = 4.5,width = 8)
```
