---
title: A closer look at some results on the 68K PBMC data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we take a closer look at some of the results on the 68K PBMC data.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load the packages used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(topicmodels)
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Load the 68K PBMC data.

```{r load-data}
load("../data/pbmc_68k.RData")
rm(counts)
```

Load the topic models fit using the EM and CD algorithms

```{r load-fits}
fit1 <- readRDS("../output/pbmc68k/rds/fit-pbmc68k-em-k=7.rds")$fit
fit2 <- readRDS("../output/pbmc68k/rds/fit-pbmc68k-scd-ex-k=7.rds")$fit
fit1 <- poisson2multinom(fit1)
fit2 <- poisson2multinom(fit2)
```

and the LDA fits initialized using the EM and CD estimates:

```{r load-lda}
lda1 <- readRDS("../output/pbmc68k/rds/lda-pbmc68k-em-k=7.rds")$lda
lda2 <- readRDS("../output/pbmc68k/rds/lda-pbmc68k-scd-ex-k=7.rds")$lda
```

The MLEs and the approximate posterior estimates from LDA turn out to
be very similar to each other, so there is really no need to examine
both. Here we'll focus on the MLEs:

```{r compare-mle-vs-lda}
cor(as.vector(fit1$L),as.vector(lda1@gamma))
cor(as.vector(fit2$L),as.vector(lda2@gamma))
```

Let's now examine the results using Structure plots. Here are the EM
estimates:

```{r structure-plot-em, fig.height=1.75, fig.width=8, results="hide", message=FALSE} 
L <- fit2$L
n <- nrow(L)
clusters <- rep("T cells",n)
names(clusters) <- rownames(fit2$L)
clusters[L[,3] > 0.4] <- "NK cells"
clusters[L[,1] > 0.3] <- "myeloid"
clusters[L[,5] > 0.2] <- "B cells"
clusters[L[,2] > 0.9] <- "Megakaryocytes"
clusters <- factor(clusters,
                   c("Megakaryocytes","myeloid","B cells","NK cells",
				     "T cells"))
n    <- nrow(fit1$L)
rows <- sample(n,2000)
fit1 <- select_loadings(fit1,rows)
fit2 <- select_loadings(fit2,rows)
k <- 7
topic_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
                  "#e31a1c","#fdbf6f")
set.seed(1)
p1 <- structure_plot(fit1,grouping = clusters[rows],topics = 1:7,
                     colors = topic_colors,gap = 20)
  ggtitle("EM without extrapolation") +
  theme(plot.title = element_text(face = "plain",size = 10))
p1
```

And here are the CD estimates:

```{r structure-plot-cd, fig.height=1.75, fig.width=8, results="hide", message=FALSE}
p2 <- structure_plot(fit2,grouping = clusters[rows],topics = 1:7,
                     colors = topic_colors,gap = 20)
  ggtitle("CD with extrapolation") +
  theme(plot.title = element_text(face = "plain",size = 10))
p2
```

```{r plots-for-paper, echo=FALSE}
ggsave("../plots/structure_plots_pbmc68k.pdf",
       plot_grid(p1,p2,nrow = 2,ncol = 1),
	   height = 3.5,width = 8)
```

The most striking difference is the greater promenance of topic 6 (red).

Let's now extract some "driving genes" for a few selected topics by
taking genes that are at higher expression in the given topic compared
to the other topics. For example, some of the top driving genes for
topic 1 suggest myeloid cells (e.g., *S100A8*, *S100A9*):

```{r driving-genes-topic-1}
k <- 1
dat <- data.frame(gene = genes$symbol,
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
dat <- transform(dat,lfc = log2(f2/f0))
subset(dat,lfc > 4 & f2 > 0.001)
```

The top driving genes for topic 5 suggest B cells (e.g., *CD79A*,
*CD19*):

```{r driving-genes-topic-5}
k <- 5
dat <- data.frame(gene = genes$symbol,
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
dat <- transform(dat,lfc = log2(f2/f0))
subset(dat,lfc > 10 & f2 > 0.0001)
```

The top driving genes for topic 3 suggest natural killer (NK) cells
(e.g., *NKG7*, *GNLY*):

```{r driving-genes-topic-3}
k <- 3
dat <- data.frame(gene = genes$symbol,
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
dat <- transform(dat,lfc = log2(f2/f0))
subset(dat,lfc > 4 & f2 > 0.001)
```

Topic 2 is less clear, but perhaps relates to megakaryocytes (e.g.,
*PF4*):

```{r driving-genes-topic-2}
k <- 2
dat <- data.frame(gene = genes$symbol,
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
dat <- transform(dat,lfc = log2(f2/f0))
subset(dat,lfc > 10 & f2 > 0.001)
```

Topic 6, the topic that is most different between the EM and CD
estimates, appears to be capturing a "baseline" expression level. This
includes ribosomal protein genes, which account for a large fraction
of the total expression in this data set, and don't capture any
"interesting" structure.

```{r driving-genes-topic-6}
k <- 6
dat <- data.frame(gene = genes$symbol,
                  f0 = apply(fit2$F[,-k],1,max),
                  f1 = fit1$F[,k],
                  f2 = fit2$F[,k])
dat <- transform(dat,lfc = log2(f2/f0),r21 = f2/f1)
subset(dat,lfc > 0.5 & f2 > 0.005)
```
