---
title: Prepare NIPS data for analysis with a topic model
author: Peter Carbonetto
output: workflowr::wflow_html
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold")
```

Here we prepare the NIPS 1--17 data for subsequent topic modeling
analyses. To run these data preparation steps, download the data from
[here][nips-data] then save the downloaded MAT file as "nips.mat"
inside the "data" directory. Note that it may be necessary to re-save
the data in MATLAB by running `save nips.mat -v6` so that it can be
successfully loaded by the `readMat` function from the R.matlab
package:


The R.matlab package is used to read the MATLAB data structures into
R.

```{r load-pkgs, warning=FALSE, message=FALSE}
library(Matrix)
library(R.matlab)
```

Documents with fewer than two nonzero counts are removed.

```{r read-data}
dat              <- readMat("../data/nips_1-17.mat")
counts           <- t(dat$counts)
rownames(counts) <- unlist(dat$docs.names)
colnames(counts) <- unlist(dat$words)
rows             <- which(rowSums(counts > 0) > 1)
counts           <- counts[rows,]
```

The word counts data should be stored as a 2,483 x 14,036
matrix. Approximately 4% of the word counts are positive.

```{r summarize-data-1}
n <- nrow(counts)
m <- ncol(counts)
cat(sprintf("Number of documents: %d\n",n))
cat(sprintf("Number of words: %d\n",m))
cat(sprintf("Rate of nonzero counts: %0.2f%%\n",100*nnzero(counts)/(n*m)))
```

Among the counts that are positive, the vast majority are small.

```{r summarize-data-2}
cat("The word counts are mostly small, with a small number of large counts:\n")
print(quantile(summary(counts)$x,c(0,0.5,0.9,0.99,0.999,1)))
```

Write the counts to an RData file.

```{r write-data}
save(list = "counts",file = "nips.RData")
```

[nips-data]: http://ai.stanford.edu/~gal/data.html
