---
title: Prepare newsgroups data for analysis with a topic model
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we prepare the "20 Newsgroups" data for subsequent topic modeling
analyses. To run these data preparation steps, download the data from
[here][20-newsgroups-data] then copy the downloaded files to a
directory called `20news-bydate` inside the `data` subdirectory.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold")
```

The Matrix package is used to create a sparse matrix.

```{r load-pkgs, warning=FALSE, message=FALSE}
library(Matrix)
library(readr)
library(tools)
```

The data were split into a "training" and "test" set. Here we combine
them into one large data set.

```{r read-data}
data.dir      <- "../data/20news-bydate"
topic.names   <- read.table(file.path(data.dir,"train.map"),sep = " ",
                            stringsAsFactors = FALSE)[[1]]
words         <- read.table(file.path(data.dir,"vocabulary.txt"),
						    stringsAsFactors = FALSE)[[1]]
topics        <- c(read.table(file.path(data.dir,"train.label"))[[1]],
                   read.table(file.path(data.dir,"test.label"))[[1]])
train         <- read.table(file.path(data.dir,"train.data"),sep = " ")
test          <- read.table(file.path(data.dir,"test.data"),sep = " ")
names(train)  <- c("document","word","count")
names(test)   <- c("document","word","count")
test$document <- test$document + max(train$document)
dat           <- rbind(train,test)
```

Create the (sparse) counts matrix.

```{r create-counts-matrix}
counts           <- sparseMatrix(i = dat$document,j = dat$word,x = dat$count)
colnames(counts) <- words
```

Remove all words that appear in fewer than 2 documents.

```{r filter-words}
cols   <- which(colSums(counts > 0) >= 2)
counts <- counts[,cols]
```

Get the "topics" assigned to the documents.

```{r get-topics}
topics         <- factor(topics)
levels(topics) <- topic.names
```

The word counts should be stored as a 18,774 x 55,911 matrix, in which
only a small proportion of the word counts (0.2%) are nonzero.

```{r summarize-data-1}
n <- nrow(counts)
m <- ncol(counts)
cat(sprintf("Number of documents: %d\n",n))
cat(sprintf("Number of terms in vocabulary: %d\n",m))
cat(sprintf("Rate of nonzero counts: %0.2f%%\n",100*nnzero(counts)/(n*m)))
```

Among the counts that are positive, the vast majority are small.

```{r summarize-data-2}
cat("The word counts are mostly small, with a small number of large counts:\n")
print(quantile(summary(counts)$x,c(0,0.5,0.9,0.99,0.999,1)))
```

Write the counts in an RData file.

```{r write-data}
save(list = c("counts","topics"),file = "newsgroups.RData")
resaveRdaFiles("newsgroups.RData")
```

[20-newsgroups-data]: http://qwone.com/~jason/20Newsgroups
