---
title: Toy example comparing EM vs. SCD
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform a small experiment with simulated data to illustrate
the behaviour of the EM and SCD algorithms for fitting Poisson NMF
(and topic models). The standard variational EM algorithm for LDA has
similar struggles in this example.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below, as well as some
additional functions used to simulate the data and generate the
results.

```{r load-pkgs, message=FALSE}
library(tm)
library(topicmodels)
library(fastTopics)
library(mvtnorm)
library(ggplot2)
library(cowplot)
source("../code/smallsim_functions.R")
```

Simulate a $100 \times 400$ counts matrix from a multinomial topic
model with $K = 6$ topics.

```{r sim-data}
set.seed(4)
n <- 100
m <- 400
k <- 6
F <- simulate_factors(m,k)
out <- simulate_loadings(n,k)
L <- out$L
major_topic <- out$major_topic
s <- simulate_sizes(n)
X <- simulate_multinom_counts(L,F,s)
cols <- which(colSums(X > 0) > 0)
F <- F[cols,]
X <- X[,cols]
```

We fit the multinomial topic model by performing 80 EM updates or 80
SCD updates. Both of the fits were first initialized by running 20 EM
updates.

```{r fit-models, message=FALSE, results="hide"}
control <- list(extrapolate = FALSE,numiter = 4,nc = 2)
fit0 <- fit_poisson_nmf(X,k,numiter=20,method="em",control=control)
fit1 <- fit_poisson_nmf(X,fit0=fit0,numiter=80,method="em",control=control)
fit2 <- fit_poisson_nmf(X,fit0=fit0,numiter=80,method="scd",control=control)
```

EM and SCD produce quite different estimates, and among the two, the
SCD estimates are clearly much closer to the ground truth.

```{r compare-fits, fig.height=4, fig.width=5, message=FALSE}
topic_colors <- c("dodgerblue","darkorange","forestgreen","darkblue",
                  "gold","skyblue")
loadings_order <- order(major_topic,L[,1])
k_set <- c(1,3,5,2,4,6)
p1 <- simdata_structure_plot(L,loadings_order,topic_colors,title = "true")
p2 <- simdata_structure_plot(poisson2multinom(fit1)$L,loadings_order,
                             topic_colors[k_set],title = "EM")
p3 <- simdata_structure_plot(poisson2multinom(fit2)$L,loadings_order,
                             topic_colors[k_set],title = "SCD")
plot_grid(p1,p2,p3,nrow = 3,ncol = 1)
```

Indeed, the SCD estimates also improve upon the EM estimates in terms
of log-likelihood, with a total improvement of about 600
log-likelihood units,

```{r compare-logliks}
make_colvec(c("em" = sum(loglik_multinom_topic_model(X,fit1)),
              "scd" = sum(loglik_multinom_topic_model(X,fit2))))
```

or an average of about 6 log-likelihood units per document,

```{r compare-logliks-2}
make_colvec(c("em" = sum(loglik_multinom_topic_model(X,fit1)),
              "scd" = sum(loglik_multinom_topic_model(X,fit2)))/n)
```

Next we will see that the variational EM (VEM) algorithm for LDA in
the [topicmodels][topicmodels] package similarly has trouble making
good progress on this data set, and greatly benefits from a good
initialization provided by the SCD algorithm. (Note that the
topicmodels package uses the [original C implementation][lda-c] of the
VEM algorithm.) To show this, we run the EM algorithm in which the
parameters are initialized to the estimates obtained by running the EM
or SCD updates:

```{r run-lda}
lda0 <- run_lda(X,fit0,numiter = 100)
lda1 <- run_lda(X,fit1,numiter = 100)
lda2 <- run_lda(X,fit2,numiter = 100)
```

Initalizing to the SCD updates indeed results in a much better fit (as
measured by the lower bound to the log-likelihood, *i.e.*, the "ELBO"):

```{r compare-elbos}
make_colvec(c("default init" = logLik(lda0),
              "100 EM init" = logLik(lda1),
              "20 EM + 80 SCD" = logLik(lda2)))
```

Note that in this example we have fixed $\alpha$, which determines the
prior on the topic-word frequencies (the F matrix in fastTopics), to 1
rather than estimate it. It is possible to let $\alpha$ adapt to the
data, but doing so complicates the comparison.

The Structure plots show that the SCD initialization leads to
estimates that look much more like the true values than the other
initializations:

```{r compare-lda-fits, fig.height=5.5, fig.width=5, message=FALSE}
p1 <- simdata_structure_plot(L,loadings_order,topic_colors,title = "true")
p2 <- simdata_structure_plot(lda0@gamma,loadings_order,
                             topic_colors[k_set],
							 title = "Initialized with 20 EM updates")
p3 <- simdata_structure_plot(lda1@gamma,loadings_order,
                             topic_colors[k_set],
							 title = "Initialized with 100 EM updates")
p4 <- simdata_structure_plot(lda2@gamma,loadings_order,
                             topic_colors[k_set],
							 title = "Initialized with 20 EM + 80 SCD updates")
plot_grid(p1,p2,p3,p4,nrow = 4,ncol = 1)
```

Let's now run the VEM algorithm for longer to see what happens.

```{r run-lda-longer}
lda0 <- run_lda(X,fit0,numiter = 800)
lda1 <- run_lda(X,fit1,numiter = 800)
lda2 <- run_lda(X,fit2,numiter = 800)
```

The VEM updates continue to approach the estimates obtained from the
SCD initialization, but they remain quite far away even after 800
iterations.

These plots show the improvement in the objective (the ELBO) over
time.

```{r elbo-plot, fig.height=2.5, fig.width=4}
pdat <- rbind(data.frame(iter = 1:800,elbo = lda0@logLiks,init = "20 EM"),
              data.frame(iter = 1:800,elbo = lda1@logLiks,init = "100 EM"),
              data.frame(iter = 1:800,elbo = lda2@logLiks,init = "20 EM + 80 SCD"))
pdat <- transform(pdat,elbo = max(elbo) - elbo + 0.01)
ggplot(pdat,aes(x = iter,y = elbo,color = init)) +
  geom_line(linewidth = 0.75) +
  scale_x_continuous(breaks = seq(0,800,100)) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-1,4)) +
  scale_color_manual(values=c("darkblue","dodgerblue","darkorange")) +
  labs(x = "iteration",y = "ELBO difference") +
  theme_cowplot(font_size = 10)
```

Returning to the maximum-likelihood estimation problem, it is
reassuring is that if we continue to perform the EM updates, they
eventually arrive at the same solution as SCD. But SCD is able to
"rescue" the EM estimates much more quickly after performing just a
few SCD updates:

```{r fit-models-2, message=FALSE, results="hide", fig.height=2.5, fig.width=3.5}
fit3 <- fit_poisson_nmf(X,fit0=fit1,numiter=700,method="em",control=control)
control$extrapolate <- TRUE
fit2 <- fit_poisson_nmf(X,fit0=fit2,numiter=700,method="scd",control=control)
fit4 <- fit_poisson_nmf(X,fit0=fit1,numiter=700,method="scd",control=control)
fit1 <- poisson2multinom(fit1)
fit2 <- poisson2multinom(fit2)
fit3 <- poisson2multinom(fit3)
fit4 <- poisson2multinom(fit4)
# loadings_scatterplot(F[,k_set],fit1$F,topic_colors,"true","em")
# loadings_scatterplot(F[,k_set],fit2$F,topic_colors,"true","scd")
pdat <- rbind(data.frame(iter   = 1:800,
			             ll     = fit2$progress$loglik.multinom,
						 method = "scd"),
              data.frame(iter   = 1:800,
                         ll     = fit3$progress$loglik.multinom,
                         method = "em"),
			  data.frame(iter   = 1:800,
			             ll     = fit4$progress$loglik.multinom,
			             method = "em+scd"))
pdat <- transform(pdat,ll = max(ll) - ll + 0.1)
ggplot(pdat,aes(x = iter,y = ll,color = method)) +
  geom_line(linewidth = 0.75) +
  scale_x_continuous(breaks = seq(0,800,100)) +
  scale_y_continuous(trans = "log10",breaks = 10^seq(-1,4)) +
  scale_color_manual(values = c("dodgerblue","darkorange","magenta")) +
  labs(x = "iteration",y = "loglik difference") +
  theme_cowplot(font_size = 10)
```

[topicmodels]: https://doi.org/10.18637/jss.v040.i13
[lda-c]: https://www.cs.columbia.edu/~blei/lda-c/
